---
title: "R Notebook"
output: html_notebook
---


---

Shaleigh Smith
NGS Final

---

Cellranger Results:
- Unfiltered: Contains every barcode from fixed list of known-good barcode sequences. This includes background and cell associated barcodes.
- Filtered: Contains only detected cellular barcodes.

---

General Steps:
1) Initialize a Seurat object
2) Filter cells based on built-in metrics
3) Normalize the data
4) Select variable genes across cells
5) Removing confounding factors by regression
6) Linear dimensionality reduction
7) PCA-based clustering
8) t-SNE visualization 
9) Identify differentially expressed genes

---

# Helpful Links:

https://github.com/igordot/seurat/blob/master/R/preprocessing.R
https://satijalab.org/seurat/v3.0/immune_alignment.html 
https://github.com/satijalab/seurat/wiki 
https://broadinstitute.github.io/2019_scWorkshop/index.html 
https://hemberg-lab.github.io/scRNA.seq.course/index.html
#https://satijalab.org/seurat/v3.0/pbmc3k_tutorial.html


---


```{r}

# Load libraries
library(tidyverse)
library(dplyr)
library(Matrix)
library(devtools)
library(Seurat)

```


---


Preprocessing

---


```{r}

# Set working directory
setwd("/Users/sha/Desktop/NGS_Informatics/NGS_final_project/")

### Preprocessing 
# Import matrices 
# Create Seurat objects 
# Set minimum for cells (3) and features (200) to remove low quality cells 

# Uninfected
data_dir_un <- "./uninfected/filtered_feature_bc_matrix/"
list.files(data_dir_un) # Should show barcodes.tsv, genes.tsv, and matrix.mtx
exp_matrix_un <- Read10X(data.dir = data_dir_un)
un <- CreateSeuratObject(counts = exp_matrix_un, project = "uninfected",
                         min.cells = 3, min.features = 200)
un$comp <- "uninfected"
un


# Latent
data_dir_lat <- "./latent/filtered_feature_bc_matrix/"
list.files(data_dir_lat) # Should show barcodes.tsv, genes.tsv, and matrix.mtx
exp_matrix_lat <- Read10X(data.dir = data_dir_lat)
lat = CreateSeuratObject(counts = exp_matrix_lat, project = "latent",
                         min.cells = 3, min.features = 200)
lat$comp <- "latent"
lat


# Ly20
data_dir_ly20 <- "./ly20/filtered_feature_bc_matrix/"
list.files(data_dir_ly20) # Should show barcodes.tsv, genes.tsv, and matrix.mtx
exp_matrix_ly20 <- Read10X(data.dir = data_dir_ly20)
ly20 = CreateSeuratObject(counts = exp_matrix_ly20, project = "ly20",
                         min.cells = 3, min.features = 200)
ly20$comp <- "ly20"
ly20


# Uninfected
data_dir_ly48 <- "./ly48/filtered_feature_bc_matrix/"
list.files(data_dir_ly48) # Should show barcodes.tsv, genes.tsv, and matrix.mtx
exp_matrix_ly48 <- Read10X(data.dir = data_dir_ly48)
ly48 = CreateSeuratObject(counts = exp_matrix_ly48, project = "ly48",
                         min.cells = 3, min.features = 200)
ly48$comp <- "ly48"
ly48 

```

```{r}

# View examples
colnames(un)
rownames(un)
head(x = un@meta.data, 20)
un[["RNA"]]@data

?seurat

```


```{r}

### Quality Control & Filtering

# View summary of total expression by single cell
summary(colSums(un))
summary(colSums(lat))
summary(colSums(ly20))
summary(colSums(ly48))

### Note that the mean decreases from uninfected to latent to ly 
### This suggests that host gene expression is decreasing as viral expression increases 


# Calculate percentage of all counts belonging to a subset of the possible features for each cell
VlnPlot(object = un, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2, 
        cols = c("#007BC3"),  pt.size = 0.25)
VlnPlot(object = lat, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2, 
        cols = c("#007BC3"),  pt.size = 0.25)
VlnPlot(object = ly20, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2, 
        cols = c("#007BC3"),  pt.size = 0.25)
VlnPlot(object = ly48, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2, 
        cols = c("#007BC3"),  pt.size = 0.25)

# Visualize feature relationships: Count_RNA vs. Feature_RNA
scatter_un_1 <- FeatureScatter(object = un, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",
                               col = "darkblue") + xlim(c(0, 85000)) + ylim(c(0, 8000))
scatter_lat_1 <- FeatureScatter(object = lat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",
                                col = "darkred") + xlim(c(0, 85000)) + ylim(c(0, 8000))
scatter_ly20_1 <- FeatureScatter(object = ly20, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",
                                 col = "darkgoldenrod2") + xlim(c(0, 85000)) + ylim(c(0, 8000))
scatter_ly48_1 <- FeatureScatter(object = ly48, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",
                                 col = "darkgreen") + xlim(c(0, 85000)) + ylim(c(0, 8000))
scatter_un_1
scatter_lat_1
scatter_ly20_1
scatter_ly48_1
#CombinePlots(plots = list(scatter_un_1, scatter_lat_1, scatter_ly20_1, scatter_ly48_1), ncol = 2)

### All appear to be a horizontal parabola (k * sqrt(x) relationship)
### All have a fairly high positive correlation (greater than 0.87 )

```


```{r}

# Distribution of expression levels by Feature and Cell
hist(rowSums(un),
     breaks = 100, main = "Uninfected: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(un),
     breaks = 100, main = "Uninfected: Expression sum by Cell",
     xlab = "Sum Expression")

hist(rowSums(lat),
     breaks = 100, main = "Latent: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(lat),
     breaks = 100, main = "Latent: Expression sum by Cell",
     xlab = "Sum Expression")

hist(rowSums(ly20),
     breaks = 100, main = "LY20: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly20),
     breaks = 100, main = "LY20: Expression sum by Cell",
     xlab = "Sum Expression")

hist(rowSums(ly48),
     breaks = 100, main = "LY48: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly48),
     breaks = 100, main = "LY48: Expression sum by Cell",
     xlab = "Sum Expression")

```



```{r}

### Normalize each gene expression matrix 
# Do Log Normalization first with default scale

# Uninfected Log Normalization
un_norm_log <- NormalizeData(un, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(un_norm_log),
     breaks = 100, main = "Uninfected Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(un_norm_log),
     breaks = 100, main = "Uninfected Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# Latent Log Normalization
lat_norm_log <- NormalizeData(lat, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(lat_norm_log),
     breaks = 100, main = "Latent Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(lat_norm_log),
     breaks = 100, main = "Latent Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# ly20 Log Normalization
ly20_norm_log <- NormalizeData(ly20, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(ly20_norm_log),
     breaks = 100, main = "LY20 Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly20_norm_log),
     breaks = 100, main = "LY20 Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# ly48 Log Normalization
ly48_norm_log <- NormalizeData(ly48, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(ly48_norm_log),
     breaks = 100, main = "LY48 Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly48_norm_log),
     breaks = 100, main = "LY48 Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

```

```{r}

### Check CLR Normalization for comparison

# Uninfected CLR Normalization
un_norm_clr <- NormalizeData(un, assay = NULL, normalization.method = "CLR", 
                         scale.factor = 10000, margin = 1)
hist(rowSums(un_norm_clr),
     breaks = 100, main = "Uninfected CLR Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(un_norm_clr),
     breaks = 100, main = "Uninfected CLR Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# Latent CLR Normalization
lat_norm_clr <- NormalizeData(lat, assay = NULL, normalization.method = "CLR", 
                         scale.factor = 10000, margin = 1)
hist(rowSums(lat_norm_clr),
     breaks = 100, main = "Latent CLR Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(lat_norm_clr),
     breaks = 100, main = "Latent CLR Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# ly20 CLR Normalization
ly20_norm_clr <- NormalizeData(ly20, assay = NULL, normalization.method = "CLR", 
                         scale.factor = 10000, margin = 1)
hist(rowSums(ly20_norm_clr),
     breaks = 100, main = "LY20 CLR Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly20_norm_clr),
     breaks = 100, main = "LY20 CLR Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# ly48 CLR Normalization
ly48_norm_clr <- NormalizeData(ly48, assay = NULL, normalization.method = "CLR", 
                         scale.factor = 10000, margin = 1)
hist(rowSums(ly48_norm_clr),
     breaks = 100, main = "LY48 CLR Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly48_norm_clr),
     breaks = 100, main = "LY48 CLR Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

```


```{r}

### Remove unwanted cells from the dataset to decrease noise after normalization

# The normalized uninfected appears to be bimodal
# To decrease the smalled leftward peak, increase the minimum number of features by cell

un_sub <- subset(x = un, subset = nFeature_RNA > 300 & nFeature_RNA > 6000)
lat_sub <- subset(x = lat, subset = nFeature_RNA > 300 & nFeature_RNA > 6000)
ly20_sub <- subset(x = ly20, subset = nFeature_RNA > 300 & nFeature_RNA > 6000)
ly48_sub <- subset(x = ly48, subset = nFeature_RNA > 300 & nFeature_RNA > 6000)


```

```{r}

### Do normalization on subsetted matrix
# Use Log Normalization
# Alter subset to increase normal distribution

### Normalize each gene expression matrix 
# Do Log Normalization first with default scale

# Uninfected Log Normalization
un_sub_norm_log <- NormalizeData(un_sub, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(un_sub_norm_log),
     breaks = 100, main = "Uninfected Sub Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(un_sub_norm_log),
     breaks = 100, main = "Uninfected Sub Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# Latent Log Normalization
lat_sub_norm_log <- NormalizeData(lat_sub, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(lat_sub_norm_log),
     breaks = 100, main = "Latent Sub Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(lat_sub_norm_log),
     breaks = 100, main = "Latent Sub Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# ly20 Log Normalization
ly20_sub_norm_log <- NormalizeData(ly20_sub, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(ly20_sub_norm_log),
     breaks = 100, main = "LY20 Sub Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly20_sub_norm_log),
     breaks = 100, main = "LY20 Sub Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

# ly48 Log Normalization
ly48_sub_norm_log <- NormalizeData(ly48_sub, assay = NULL, normalization.method = "LogNormalize", 
                         scale.factor = 10000)

hist(rowSums(ly48_sub_norm_log),
     breaks = 100, main = "LY48 Sub Log Normalized: Expression sum by Feature",
     xlab = "Sum Expression")
hist(colSums(ly48_sub_norm_log),
     breaks = 100, main = "LY48 Sub Log Normalized: Expression sum by Cell",
     xlab = "Sum Expression")

```

---

```{r}

# Integrate data for comparisons (uninfected vs latent, latent vs ly20, ly20 vs ly48)

# Ex.
immune.anchors <- FindIntegrationAnchors(object.list = list(ctrl, stim), dims = 1:20)
immune.combined <- IntegrateData(anchorset = immune.anchors, dims = 1:20)

# https://satijalab.org/seurat/v3.0/immune_alignment.html 

```


---


---